### 2019-05-02 papers/haskell19/5-typelevel.tex
AddedAle's feedback

% --------------------------------------
% Intro
%
This section introduces a set refinements to our basic language for describing
random generators, making them more flexible and robust in order to fit
real-world usage scenarios.

% --------------------------------------
% Naming conventions
%
To begin with, let us consider the name conventions we used when defining our
representation data types along this paper.
%
Since our aim is to automate this process as much as possible using Haskell's
meta-programming capabilities, the actual names used when deriving construction
representations will have to be generated in such a way that they are compliant
with Haskell syntax and are unique within our namespace.
%
This means that, in practice, type names like |Fun_cat| or |Pat_simplify_1| are
technically not valid Haskell data type names, thus they will have to be
synthesized as something like |Fun_lt_plus_gt_543| and |Pat_simplify_1_325|,
where the last sequence of numbers is automatically inserted by Template Haskell
to ensure uniqueness.

This naming convention results hard to use, specially if we consider that we do
not now the actual name types until they are synthesized in compile-time, due to
their unique suffixes.
%
Fortunately, it is easy to solve this problem again using some type-level
machinery.
%
Instead of imposing a naming convention in our representation synthesizer, we
can define a set of open type families to hide each kind of construction using
meaningful names:

@@

Where |Symbol| is the kind of type-level strings in Haskell.
%
Then, our derivation process will synthesize each representation using unique
names, along with a type instance of the corresponding type family for the end
user to manipulate it.
%
For instance, along with the representations of |Con_Text|, |Fun_cat| and
|Pat_simplify_1|, we will automatically derive the following type instances:

@@

And the end user can simply refer to each particular construction by using these
synonyms, e.g., with representation types like |Con "Text" oplus' Fun "<+>"|.

@@

This way, any quoted construction will considered to be terminal.
%
With these type-level manipulations, we are now able to write generation
specifications in a clear and concise way.
%
For instance, we can specify a complex random generator for |Html| values as
follows:

@@

And simply obtain a concrete generator by refering to this previous type when
using |genEval|:

@@

So far we have seen how specify random generators for our simple self-contained
|Html| data type.
%
In practice, however, we are often required to write random generators for
parametric target data types as well.
%
Consider for example the following |Tree| data type definition encoding binary
trees with generic information of type |a| in the leaves:

@@

\begin{code}
data Con_Leaf a r = Con_Leaf a
data Con_Node a r = Con_Node r r
\end{code}

@@

\begin{code}
instance Algebra (Con_Leaf a) (Tree a) where ...
instance Algebra (Con_Node a) (Tree a) where ...

instance Arbitrary a => BArbitrary1 (Con_Leaf a) where ...
instance Arbitrary a => BArbitrary1 (Con_Node a) where ...
\end{code}

The biggest different appears when deriving the required type instances for the
|Con|, |Fun| and |Pat| synonyms.
%
Instead of carrying this type parameter in our generation specifications, we can
avoid it by hiding them using an existential type:

\begin{code}
data Some (f :: * -> * -> *) (r :: *) = forall (a :: *) dot Some (f a r)
\end{code}

In the previous definition, |Some| is a wrapper for a 2-parametric type that
hides the first type variable by using an explicit existential quantificator.
%
Note thus that the type parameter |a| does not appear at the left hand side of
|Some| in its defition.
%
Then, when deriving any |Con|, |Fun| or |Pat| type synonym, we will wrap each
construction representations using a |Some|:

@@

This way we can write generation specification for our |Tree| data type without
having to refer to its type parameter anywhere.
%
For instance:

@@

\begin{code}
type family (f :: * -> *) apply' (a :: *) :: * -> * space where
  (Some f)         apply'  a  = f a
  (f oplus''  g)   apply'  a  = (f apply' a) oplus'' (g apply' a)
  (f otimes'' n)   apply'  a  = (f apply' a) otimes'' n
  (Term f)         apply'  a  = Term (t apply' a)
  f                apply'  a  = f
\end{code}

As a result, in our definition of |genIntTree| above, the |apply| operator will
reduce the type |(Tree_Spec apply'' Int)| to the following concrete type:

@@

Along with our automated constructions derivation mechanism, the machinery
introduced in this section allows us to specify random generators using a
concise and flexible type-level domain specific language.
%
The next section evaluates our approach in terms of performance using a set of
case studies extracted from real-world Haskell libraries.

>>>>

% --------------------------------------
% Intro
%
This section introduces refinements to our basic language for describing random
generators, making them more flexible and robust in order to fit real-world
usage scenarios.

% --------------------------------------
% Naming conventions
%
% To begin with, let us consider the name conventions we used when defining our
% representation data types along this paper.
%
% Since our aim is to automate this process as much as possible using Haskell's
% meta-programming capabilities,
%
The first problem we face is that of naming conventions.
%
In practice, the actual names used when deriving construction representations
need to be generated in a way that they are compliant with Haskell syntax and
are \emph{unique within our namespace}.
%
This means that, type names like |Fun_cat| or |Pat_simplify_1| are technically
not valid Haskell data type names, thus they will have to be synthesized as
something like |Fun_lt_plus_gt_543| and |Pat_simplify_1_325|, where the last
sequence of numbers is automatically inserted by Template Haskell to ensure
uniqueness.

This naming convention results hard to use, specially if we consider that we do
not know the actual type names until they are synthesized in compile-time, due
to their unique suffixes.
%
Fortunately, it is easy to solve this problem using some type-level machinery.
%
Instead of imposing a naming convention in our tool, we can define a set of open
type families to hide each kind of construction using meaningful names:

@@

where |Symbol| is the kind of type-level strings in Haskell.
%
Then, our derivation process will synthesize each representation using unique
names, along with a type instance of the corresponding type family for the end
user to manipulate it.
%
For instance, along with the representations of |Con_Text|, |Fun_cat| and
|Pat_simplify_1|, we will automatically derive the following type instances:

@@

As a result, the end user can simply refer to each particular construction by
using these synonyms, e.g., with representation types like |Con "Text" oplus'
Fun "<+>"|.

@@

This way, any construction with the supra-index |T| will considered to be
terminal.
%
With these type-level manipulations, we are now able to write generation
specifications in a clear and concise way.
%
For instance, we can specify a complex random generator for |Html| values as
follows:

@@

And simply obtain a concrete generator by refering to |Html_Spec| when using
|genEval|:

@@

So far we have seen how to specify random generators for our simple
self-contained |Html| data type.
%
In practice, however, we are often required to write random generators for
parametric target data types as well.
%
Consider, for example, the following |Tree| data type definition encoding binary
trees with generic information of type |a| in the leaves:

@@

\begin{code}
data Con_Leaf a r = Mk_Leaf a
data Con_Node a r = Mk_Node r r
\end{code}

@@

\begin{code}
instance Algebra (Con_Leaf a) (Tree a) where dots
instance Algebra (Con_Node a) (Tree a) where dots

instance Arbitrary a => BArbitrary1 (Con_Leaf a) where dots
instance Arbitrary a => BArbitrary1 (Con_Node a) where dots
\end{code}

% The biggest different appears when deriving the required type instances for the
% |Con|, |Fun| and |Pat| synonyms.
%
Instead of carrying this type parameter in our generation specifications, we can
avoid it by hiding them using an existential type:

\begin{code}
data Some (f :: * -> * -> *) (r :: *) = forall (a :: *) dot SomeC (f a r)
\end{code}

The type constructor |Some| is a wrapper for a 2-parametric type that hides the
first type variable using an explicit existential quantificator.
%
Note thus that the type parameter |a| does not appear at the left hand side of
|Some| in its defition.
%
Then, when deriving any |Con|, |Fun| or |Pat| type synonym, we will wrap each
construction representations using a |Some|:

@@

We can therefore write generation specification for our |Tree| data type without
having to refer to its type parameter anywhere.
%
For instance:

@@

\begin{code}
type family (f :: * -> *) apply' (a :: *) :: * -> * space where
  (Some f)         apply'  a  = f a
  (f oplus''  g)   apply'  a  = (f apply' a) oplus'' (g apply' a)
  (f otimes'' n)   apply'  a  = (f apply' a) otimes'' n
  (Term f)         apply'  a  = Term (t apply' a)
  f                apply'  a  = f
\end{code}
%
As a result, in |genIntTree|, the |apply| operator will reduce the type
|(Tree_Spec apply'' Int)| to the following concrete type:

@@

Along with our automated constructions derivation mechanism, the machinery
introduced in this section allows us to specify random generators using a
concise and flexible type-level domain specific language.
%
The next section evaluates our approach in terms of performance using a set of
case studies extracted from real-world Haskell implementations.

### 2019-05-02 papers/haskell19/6-casestudies.tex
AddedAle's feedback

The random generation framework presented throughout this work allows us to
write extensible generators in a very concise way.
%
However, this expresiveness comes attached to a perceptible runtime overhead.
%
In this section we evaluate the implicit cost of composing generators using
three real-world case studies.

@@

To evaluate our approach, we analyzed the performance of generating random
values using three real-world case studies:
\begin{inparaenum}[(i)]
\item Red-Black Trees, inspired by Okasaki's formulation,
\item Lisp S-expressions, inspired by the package \emph{hs-zuramaru}, and
\item HTML expressions, inspired by the \emph{html} package, and following the
  same structure as our motivating |Html| example.
\end{inparaenum}
%
The expanse of each case study can be summarized as follows:

\begin{table}[H]
  \begin{tabular}{||c||c||c||c||c||}
    \hline
    Case Study & |Con|s & |Pat|s & |Fun|s & Total Constructions \\ \hline
    \hline
    RBT        & 2      & 6      & 5      & 13                  \\ \hline
    SExp       & 6      & 9      & -      & 15                  \\ \hline
    HTML       & 4      & -      & 132    & 136                 \\ \hline
  \end{tabular}
\end{table}

>>>>

The random generation framework presented throughout this work allows us to
write extensible generators in a very concise way.
%
However, this expresiveness comes attached to a perceptible runtime overhead,
primarly inherited from the using the \`a la Carte technique.
%
In this section we evaluate the implicit cost of composing generators using
three real-world case studies.

@@

In this work, we analyzed the performance of generating random values using
three case studies:
%
\begin{inparaenum}[(i)]
\item Red-Black Trees, inspired by Okasaki's formulation,
\item Lisp S-expressions, inspired by the package \emph{hs-zuramaru}, and
\item HTML expressions, inspired by the \emph{html} package, which follows the
  same structure as our motivating |Html| example.
\end{inparaenum}
%
The expanse of each case study can be summarized as follows:

\begin{table}[H]
  \begin{tabular}{||c||c||c||c||c||}
    \hline
    Case Study & \#|Con| & \#|Pat| & \#|Fun| & Total Constructions \\ \hline
    \hline
    RBT        & 2      & 6      & 5      & 13                  \\ \hline
    SExp       & 6      & 9      & -      & 15                  \\ \hline
    HTML       & 4      & -      & 132    & 136                 \\ \hline
  \end{tabular}
\end{table}

### 2019-05-02 papers/haskell19/7-related.tex
AddedAle's feedback

\emph{DRAGEN} is a random testing derivation tool that synthesizes random
generators for our target data types, tuning their generation frequencies using
a simulation-based optimization process parametrized by the distribution of
values desired by the user \cite{DBLP:conf/haskell/MistaRH18}.
%
This simulation is based on the theory of \emph{branching processes}, which
models the growth and extinction of populations across succesive generations.
%
In this setting, populations consist of randomly generated data constructors,
where generators correspond to each level of the generated values.
%
This tool has shown to improve the code coverage obtained over our codebase,
with respect to other fully-automated generators derivation mechanism.
%
Recently, \citeauthor{Mista2019GeneratingRS} extended this approach to generate
random values considering also the other sources of structural information
covered here, namely abstract interfaces and function pattern machings.
%
Using this extension, the authors show how the usage of extra information when
generating random values is extremely valuable, particularly in situations like
the ones described in Section \ref{sec:sources}, where the usual derivation
approaches fail to synthesize useful generators due to a lack of structural
information.
%
Our work is inspired by the same general idea. In constrast, in this paper we
tackle the representation problem, showing how these ideas can be effectively
implemented in Haskell using advanced type-level features.

>>>>

\emph{DRAGEN} is a random testing derivation tool that synthesizes random
generators for our target data types, tuning their generation frequencies using
a simulation-based optimization process parametrized by the distribution of
values desired by the user \cite{DBLP:conf/haskell/MistaRH18}.
%
This simulation is based on the theory of \emph{branching processes}, which
models the growth and extinction of populations across succesive generations.
%
In this setting, populations consist of randomly generated data constructors,
where generations correspond to each level of the generated values.
%
This tool has shown to improve the code coverage obtained over our codebase,
with respect to other fully-automated generators derivation mechanism.
%
Recently, \citeauthor{Mista2019GeneratingRS} extended this approach to generate
random values considering also the other sources of structural information
covered here, namely abstract interfaces and function pattern machings.
%
Using this extension, the authors show how the usage of extra information when
generating random values is extremely valuable, particularly in situations like
the ones described in Section \ref{sec:sources}, where the usual derivation
approaches fail to synthesize useful generators due to a lack of structural
information.
%
Our work is inspired by the same general idea. In constrast, this paper tackles
the representation problem, showing how these ideas can be effectively
implemented in Haskell using advanced type-level features.
